Demystifying the Computer Science PhD Admission in the US\\{\large A Guide for International Students

### 2024-01-02 demystify.tex
updates

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the PhD admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this document is primarily intended for students interested in CS, it may be relevant to students from various disciplines. 
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) universities are classified as R1.} institutions in the US (and also likely universities in other countries).

@@

The latest version of this document is at \url{nguyenthanhvuh.github.io/phd-cs-us/demystify.pdf} and its \LaTeX{} source is on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{Github}. If you have questions or comments, feel free to create a \href{https://github.com/nguyenthanhvuh/phd-cs-us/issues}{GitHub issue} for discussion.

@@

Here, we outline the main points of the document. This gives you an overview to decide which specific topics you want to explore more thoroughly.

@@

Second, applying to a good US university \emph{should not} be any harder than at schools in other countries. In fact, it might even be more flexible since CS PhD in the US \emph{do not} require having an MS or knowing a topic/proposal/adviser in advance (by the way, GRE is also not required!). If you believe you have a chance in other countries, e.g., South Korea, Singapore, Germany, UK, Japan, and Australia, then you will surely have a chance in the US as well. \S\ref{sec:non-us-differences} compares CS PhD study in the US to other countries.

@@

\begin{domesticbox}[Domestic students:] If you're a domestic student, you have several advantages in your application.  First, we already know about your school (see \S\ref{sec:your-school}), and you are also more familiar with the US education system and culture.  You also have more opportunities for funding, e.g., through government scholarships for US citizens and residents. Finally, compared to international students, PhD applications in CS from domestic students are few and that can help your case.
\end{domesticbox}

@@

\section*{References}
\begin{itemize}
    \item \href{https://parentheticallyspeaking.org/articles/us-cs-phd-faq/}{Getting a Computer Science PhD in the USA} by Shriram Krishnamurthi
\end{itemize}

@@

After you submit your PhD application (usually November--December), it will be first screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters?

Then, applications are reviewed by a \textbf{PhD admission committee} that consists of faculty members in CS (in some cases the committee can involve affiliated faculty from different disciplines). These faculty have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the committee depends on the department size. At GMU, the PhD committee typically has 15--20 faculty, and each committee member is assigned to review 25--30 applications. Note that GMU, UMass, and probably many other schools, have separate committees for MS programs.

@@

The PhD admission committee typically involves assistant professors in the department (see \S\ref{sec:faculty-types} for various faculty titles). This provides junior faculty the opportunities to recruit students. The chair of the committee will be a senior professor, but they likely will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty.

@@

\begin{commentbox}[Hakan:]
  At GMU, for full consideration, students should make sure to submit \textbf{ALL} required documents by the application deadline, and should never assume that some required documents (such as official TOEFL scores or official diplomas/transcripts) will be waived by the admissions office. If something is listed and not marked as ``optional'', it is mandatory and they should plan for submitting all those.
\end{commentbox}
\paragraph{Why we do not waive application fee?}  This is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

>>>>

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the PhD admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this document is primarily intended for students interested in CS, it may be relevant to students from various disciplines. 
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) universities are classified as R1.} institutions in the US (and universities in other countries).

@@

This document is available at \url{nguyenthanhvuh.github.io/phd-cs-us/demystify.pdf} and its \LaTeX{} source is on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{Github}. If you have questions or comments, feel free to create a \href{https://github.com/nguyenthanhvuh/phd-cs-us/issues}{GitHub issue} for discussion.

@@

The main points of this guideline are summarized here. This gives you an overview to decide which specific topics you want to explore more thoroughly.

@@

Second, applying to a good US university \emph{should not} be any harder than at schools in other countries. In fact, it might even be more flexible since CS PhD in the US \emph{do not} require having an MS or knowing a topic/proposal/adviser in advance (by the way, GRE is not required!). If you believe you have a chance in other countries, e.g., South Korea, Singapore, Germany, UK, Japan, and Australia, then you will surely have a chance in the US as well. \S\ref{sec:non-us-differences} compares CS PhD study in the US to other countries.

@@

\begin{domesticbox}[Domestic students:] If you're a domestic student, you have several advantages in your application.  First, we already know about your school (see \S\ref{sec:your-school}), and you are also more familiar with the US education system and culture.  You also have more opportunities for funding, e.g., through government scholarships for US citizens and residents. Finally, compared to international students, PhD applications in CS from domestic students are few and that can really help your case.
\end{domesticbox}

@@

\section*{Additional Resources}
\begin{itemize}
    \item \href{https://parentheticallyspeaking.org/articles/us-cs-phd-faq/}{Getting a Computer Science PhD in the USA} by Shriram Krishnamurthi
\end{itemize}

@@

After you submit your PhD application (usually November--December), it will first be screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters? Usually, this screen process is done through a central university system (i.e., not by the CS department).  If you pass this screen, your application will be forwarded to the CS department for further evaluation. If you don't, then the system will inform you what is missing and what you need to do.

\begin{commentbox}[Hakan:]
    At GMU, for full consideration, students should make sure to submit \textbf{ALL} required documents by the application deadline, and should never assume that some required documents (such as official TOEFL scores or official diplomas/transcripts) will be waived by the admissions office. If something is listed and not marked as ``optional'', it is mandatory and they should plan for submitting all those.
  \end{commentbox}

Then, applications are reviewed by a \textbf{PhD admission committee} (adcom) that consists of faculty members in CS (in some cases the committee can involve affiliated faculty from different disciplines). These faculty have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depends on the department size. At GMU, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review 25--30 applications. Note that GMU, UMass, and probably many other schools, have separate adcom for MS programs.

@@

The PhD adcom typically involves assistant professors in the department (see \S\ref{sec:faculty-types} for various faculty titles). This provides junior faculty the opportunities to recruit students. The chair of the committee will be a senior professor, but they likely will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty.

@@

\paragraph{Why we do not waive application fee?}  This is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.
